Revealing Benefits, Costs, and Tradeoffs of Spatial Management in Marine Systems Arrising from Coupled Economic and Ecological Dynamics

### 2022-12-23 03_marlin_paper.qmd
now with scale-free CTMC and some edits (#7)

sci_to_com <- taxize::sci2comm(unique(marlin_inputs$scientific_name),simplify= TRUE,db='ncbi') %>% 
  bind_rows(.id = "critter") %>% 
  pivot_longer(everything(),names_to = "critter",values_to = "common_name") %>% 
  mutate(common_name = stringr::str_to_title(common_name))

results_files <- list.files(file.path(results_path))

@@

`marlin` follows a "models of intermediate complexity" (MICE) philosophy
[@plagányi2014] targeted specifically towards capturing heterogeneous
habitats, movement dynamics, and the simultaneous impacts of fishing
fleets across multiple species (frequently called "technical
interactions"). In its general form, `marlin` simulates the behavior of
populations of spatially explicit biologically independent
age-structured animal populations affected by one or more fishing fleets
in time steps specified by the user (generally annual or seasonal).
While `marlin` does not simulate species interactions, fishing fleets in
`marlin` are capable of targeting and affecting multiple species
simultaneously. This allows the model to simulate processes such as
fisheries bycatch or effort displacement in a way that accounts for how
fleet behavior might affect multiple species in the system. The ability
to track the impacts of fishing fleets across multiple species
simultaneously is particularly important as very few fishing fleets are
truly single species.

As a demonstration of the use of this modeling two, we explore two
general situations

1.  Trade-offs and distributional outcomes for food production and
    multi-species biomass in a coral-reef fishery
2.  Implications of climate-driven range shifts for MPA design in a
    pelagic fishery

We chose these two examples to illustrate the use of the model in
contrasting systems in which MPAs are increasingly considered.

Below we provide a summary of the `marlin` model, as well as details of
the generalized simulation exercises and case study analyses. We note
that `marlin` is a complex software package with many options to control
detailed behaviors. Given that these specific features will evolve over
time, we do not describe in detail every option of the model here.
Rather, we focus on the equations and methods describing the core
foundations of our tool for efficiently simulating bio-economic dynamics
in aquatic systems.

While users can adapt the general equations and ideas presented here as
they see fit, the software package accompanying this paper,
[`marlin`](https://danovando.github.io/marlin/), is written in R
[@rcoreteam2021] and C++, integrated through the `Rcpp` package
@eddelbuettel2018 .

@@

`marlin` simulates the dynamics of one or more species, currently best
representing fish-like species, using age-structured population
dynamics. Each time step, species move throughout the simulated seascape
using both diffusion and "taxis" (active movement towards preferred
habitat), experiences natural and potentially fishing mortality, and if
in a spawning season spawn using one of the possible forms of density
dependence implemented in the model, with the potential to include
auto-correlated stochastic deviations in the amount of offspring
produced, generally called "recruits".

These species are caught by fishing fleets. A fishing fleet is defined
in `marlin` by a set of fishers that can have different fishing skill,
prices, and selectivity for individual species (each of which we denote
as a *metier*). For example, both a longline and purse seine fleet may
capture bigeye (*Thunnus obesus*) and skipjack tunas, but the longline
fleet can be made much more likely to catch bigeye than skipjack, and
*vice versa*. Each fishing fleet distributes its fishing effort in space
according to a specified spatial allocation rule, for example in
proportion to profit-per-unit-effort, conditional on the presence of any
spatial restrictions such as MPAs.

Unconstrained by management, the total amount of effort exerted by each
fleet can follow one of two dynamics: open-access or constant-effort.
Under constant effort, the total amount of effort of each fleet is fixed
over time, with the possible exception of attrition due to MPA
placement. Under open-access, the total amount of effort in a given time
step is a function of the profitability of the fleet in the previous
time step, evolving until a bionomic equilibrium of zero total profits
is reached. Profitability is a function of the volume and price of each
species caught, as well as the cost of the total amount of fishing
effort per fleet and the distance from port that the fleet operates at.

The dynamics of each fleet can be modified by management in a variety of
ways. Managers can impose size limits for individual species within each
fleet. They can also impose total catch quotas for each species in each
time step. When quotas are activated, if the total catch across all
fleets for a given species under the model's effort dynamics would
exceed the allowable quota for that species, the total effort for each
fleet is decreased proportional to its contribution to the total catch
until the quota is satisfied. As an alternative to catch quotas,
managers can set an effort cap per fleet, which prevents effort from
exceeding a given amount under open-access dynamics, though effort may
be reduced below this cap if required by the profit equation (i.e. the
fleet can choose to fish less than the quota or effort cap). `marlin`
also allows users to specify closed fishing seasons for one or more
species in the system. During closed seasons, no effort is allocated to
patches that would result in capture of the covered species, potentially
resulting in zero effort if the species cannot be avoided. Under these
conditions, the amount of effort in the next open season is a function
of the effort and profits in the last open fishing season. Lastly,
managers can specify locations of no-take MPAs, which could in theory
change in size and location. When MPAs are implemented, fishing effort
that used to operate inside the MPAs can either leave the fishery, or be
redistributed outside the MPA (the default behavior).

Users can use the model to explore general properties of spatially
explicit social-ecological systems, or can parameterize it to explore
key uncertainties of more tactical applications. As an example of the
later case, we provide two case studies here designed to reflect the
kinds of questions stakeholders might face if designing an MPA network
for a nearshore coral reef or a pelagic system.

@@

Revenues *R* for fleet *f* in time step *t* are a function of the total
amount caught of each species (*Y*) and its price ($\Pi$). The amount
caught is a function of the selectivity at age of each species for each
fleet ($\alpha$), the fishing efficiency of the fleet for that that
species (also called "catchability", *q*), the amount of fishing effort
of the fleet in question in that patch in a given time step (*E*), and
the total fishing mortality at age (including all other fleets) for that
particular species in that patch and time (*u*).

@@

We approximate the dome-shaped form as a simple normal distribution with
mean $l^{sel}$ and standard deviation $\sigma$. The normal density
function is re-scaled such that the selectivity is 1 at $l^{sel}$.

@@

Users can tune their fleets in one of two ways. First, they can specify
a target exploitation rate *u* for each species in their model run.
Taking all the other parameters of the model as given, `marlin` then
adjusts the catchability coefficients $q_{f,s}$ for each fleet *f* and
species *s* such that the equilibrium exploitation rate for each species
matches the desired level. Second, they can specify a target depletion,
defined as total spawning stock biomass under fishing divided by total
unfished spawning stock biomass, and the model will adjust the
catchability coefficients $q_{f,s}$ for each fleet *f* and species *s*
such that the equilibrium depletion for each species matches the desired
level.

@@

Each fishing fleet decides how to allocate its effort in space based on
one of four possible spatial allocation strategies. The ideal free
distribution (IFD) is the standard method for distributing fishing
fleets in spaces (see @gillis2003 and references therein). However,
analytical solutions to the IFD present a number of complications for
our model. Primarily, our model is capable of tracking multiple fleets
at once. This means that the IFD for a single fleet would be a
game-theoretic outcome of the actions of each of the individual fleet
conditional on the actions of all other fleets. While possible to
implement this would slow down our model runs to the point of
practically preventing large-scale evaluation of the effects of MPAs on
bycatch.

@@

Revenue based spatial distribution is not likely to be very realistic;
in general we would expect fishing fleets to respond to profits on some
core level. However, due to the complexity of parameterizing cost
functions fleet dynamics are often evaluated based on yield or revenue
alone, and so we include those scenarios here to allow users to evaluate
the potential implications of this choice. The decision on whether to
allocate the fleet based on absolute or relative (per unit effort)
metrics is more complex. When effort represents the actions of separate
and individual fishing actors (e.g. independent fishing vessels), a
per-unit-effort strategy, in which fishers distribute themselves based
on the expected catch of their individual efforts, may be more realistic
[@hilborn1987]. Conversely, a system defined by a sole owner seeking to
maximize total profits might be better represented by a fleet model
based on total profits

@@

The underlying population model used is an age structured single-species
model in the manner of @ovando2021a. The population model requires many
parameters. However, if the user supplies either a scientific or common
name for the species in question (scientific preferred), the model will
supply default values for that species based on the values reported in
FishLife [@thorson2020]. All of the life history values can
alternatively be replaced with user-supplied values as needed; the
default values simply ensure that reasonable and internally consistent
values for the species in question are supplied as a starting point.

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*diffusion*, *advection* (follow currents), and *taxis* (active movement
towards say a preferred habitat). For now, our model focuses just on the
diffusion and taxis components of this model, though future extensions
could easily incorporate advection vectors from oceanographic models. In
this way, `marlin` allows users to run anything from a simple Gaussian
dispersal kernel up to a system governed by species that passively
diffuse around a dynamic thermal range.

We provide a brief overview of the the general CTMC method here, see
@thorson2021a for a detailed description. Under this framework, movement
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of the sum of the diffusion and taxis matrices scaled by
the width of the time step that the model is running under.

$$
\boldsymbol{M}_{t,s,a} = e^{\Delta t(\boldsymbol{D}_{t,s,a} + \boldsymbol{X}_{t,s,a})} 
$$ {#eq-movement}

Where $\boldsymbol{D}$ is the diffusion matrix and $\boldsymbol{X}$ is
the taxis matrix. The diffusion matrix is a function of a user specified
diffusion rate [@thorson2021a].

We slightly modify the taxis matrix, denoted as $\tau$ in the following
manner

$$
\tau_{p1,p2,t,s,a} = \begin{cases}
      = (1 + \lambda d_{s,f})e^{h(p2,t,s,a) - h(p1,t,s,a)} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} (1 + \lambda d_{s,f})e^{h(p2,t,s,a) - h(p1,t,s,a)} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
  \end{cases}
$$ {#eq-taxis}

The $\lambda$ term is a scalar in which the magnitude of the habitat
gradient between patches can be scaled relative to the diffusion
parameter $d$. When the diffusion and taxis matrices are estimated
jointly from actual data in the manner of @thorson2021a, $\lambda$
should be set to zero. However, when users are manually creating both
the diffusion and taxis matrices, the relative strength of diffusion to
taxis must be defined by the user. Setting higher values of $\lambda$
allows users to simulate a scenario where taxis is much more significant
than diffusion.

For both the diffusion and taxis matrix, we allow for the inclusion of
physical barriers to movement (i.e. land). Pairs of patches that are
adjacent but in which one or both patches are a barrier to movement are
set as non-adjacent. The CTMC model then produces movement dynamics that
move around barriers rather than over them. This is different than
setting habitat in a patch to zero; an animal will preferentially move
towards a patch with zero habitat all else being equal, but can move
over a zero habitat patch towards another better patch.

@@

One challenge is that the coefficients of the model matrices inside
@eq-movement can be hard to interpret. To facilitate this, `marlin`
allows users to specify a maximum number of cells that adults or
recruits diffuse in a year. The model will then tune the diffusion
parameter and habitat gradients to achieve this desired level of
movement.

@@

We include two examples demonstrating how `marlin` can be used to
support marine spatial planning. In the first, we show how `marlin` can
be used to compare the total and distributional impacts of MPAs designed
in a heavily fished coral reef ecosystem. In the second, we demonstrate
how `marlin` can be used to assess components of climate resilience of
alternative MPA design strategies in a pelagic system.

@@

Across all of our case studies we make use of four potential "rule of
thumb" MPA design strategies (xx add citation here) :

@@

2.  Avoid_fishing (*avoid_fishing*): MPAs locations are prioritized
    inversely to fishery catches. So, patches with low catches are
    prioritized over patches with high catches.

3.  Target_fishing (*target_fishing*): MPAs locations are prioritized
    proportionally to fishery catches. So, patches with high total
    catches are prioritized over patches with low catches.

4.  Area target (*area*): For the "area" strategy, we simply assign
    patches in sequential order until the desired total MPA size is
    reached.

@@

We use `marlin` to simulate the outcomes for both food production and
conservation for each of the species and both of the fleets as a
function of both MPA size and MPA design strategy. For this exercise,
MPAs are placed with perfect information and have no design constraints
for continuity (i.e. the selected MPA design can be as patchwork as
dictated by the data).

@@

For this exercise, we focused on using `marlin` to assess resilience of
the selected "rule of thumb" MPA design strategies to a climate-driven
range shift. Specifically, we simulate an extreme example where the
centroid of each population shifts northward at a rate of 0.2 patches
per year, over a `r experiment_years` year time horizon (@fig-blue-ssb).
We designed MPA networks given the conditions in the starting year, and
then held that network constant over the years of the experiment,
running one simulation with and another without the climate-drive range
shift. We then compared the effects of this range shift on food
production and conservation outcomes from MPA networks designed based on
the pre-range shift world.

@@

For the nearshore coral reef example, both the *area* and
*target_fishing* MPA design strategies were capable of producing a
Pareto-improving, i.e. potentially win-win, outcome where the total
spawning biomass divided by unfished spawning biomass across all species
increase by up to 200% while maintaining at least the same levels of
fisheries yields as would be caught in the absence of any MPAs.
Conversely, achieving increases in total fish biomass through the *rate*
or *avoid_fishing* strategies resulted in immediate losses in total
yields, and in losses to both yields and biomass in some cases
(@fig-coral-results).

These macro-level changes in total yield and conservation outcomes mask
some internal tradeoffs. Nearly all of the food production benefits from
this MPA network accrued to fleet one, while catches by fleet two were
negatively impacted under nearly all scenarios, including those that
resulted in the highest total gains to fleet one and subsequently to the
system as a whole. For example, the MPA sizes and designs that result in
the largest increase in yields for fleet one would produce a near total
loss of yields for fleet two.

Both the *rate* and *avoid_fishing* MPA placement strategies produced
lower total fishery gains but minimized the negative impacts on fleet
two. However, these strategies produced much lower conservation gains
than the *area* and *target_fishing* strategies, only benefiting
deep_snapper meaningfully, and in fact resulting in slight losses to the
nearshore snapper population for MPA sizes less than roughly 75% of the
simulated space. Both the *area* and *target_fishing* strategies
resulted in rapid conservation gains across all of the simulated
species.

MPAs were capable of producing positive outcomes in terms of total
biomass and yields in the nearshore coral reef example, but all the
tested strategies resulted in tradeoffs. `marlin` could be further used
by a community such as the one simulated here to explore new design
strategies, or to assess options for distributing fishery benefits
across communities that may experience very different outcomes from the
same MPA network.

```{r}
#| label: fig-coral-results
#| fig-cap: "Frontier in conservation and yield outcomes under different MPA design strategies (A) for nearshore coral reef case study. Change in yield for each fleet (B) and depletion per species (C) as a function of MPA size and MPA design strategy for the coral reef case study."
#| fig-height: 4
#| fig-width: 6

@@

The yield impacts of the proposed MPA strategies were much more
sensitive to climate impacts for the purse seine fishery. Under the
*status quo* species distribution, the target_fishing MPA design
strategy produced increases in fisheries yields at sizes between 25-50%
of the simulated area; that same MPA network resulted in the near
complete loss of yields from the purse seine fleet when applied to the
climate change scenario. In contrast, the yield outcomes for the
longline fleet were relatively insensitive to the simulated range shifts
(@fig-blue-results).

From a conservation perspective, species such as albacore and swordfish
were relatively insensitive to the simulated range shifts, whereas
species such as bigeye tuna and silky shark showed extreme variation in
outcomes depending on the climate scenario. For bigeye tuna, under the
*status quo* conditions the target fishing MPA design strategy would
result in rapid increases in the population, but that same MPA network
would produce much smaller and slower gains under climate change.
However, for blue sharks this trend is inverted, with more rapid
conservation gains under the range shift scenario than the *status quo*
(@fig-blue-results).

@@

Marine ecosystems are driven by complex social-ecological dynamics.
Communities must make decisions on how to manage these systems based on
limited empirical evidence. Modeling tools such as `marlin` can help
users answer important scientific questions and design marine management
policies informed by a better understanding of sensitivities to key
uncertainties.

@@

In this paper we have documented the core system of equations underlying
`marlin`, and demonstrated how `marlin` can be used to support efforts
such as MPA planning. The variation in both the magnitude and
distribution of costs and benefits stemming from alternative design
choices and states of nature shown in our case study results
demonstrates the value in our method. For example, our coastal coral
reef example illustrates both the potential for MPAs to benefit
conservation and food production in these systems, and the potential for
the same MPA to benefit some fleets while harming others when fishing
fleets affected by an MPA do not share the same objectives. Our pelagic
case study illustrates how `marlin` can be used to assess the climate
resiliency of spatial management strategies, and identify strategies
that best meet the needs of both current and future conditions. Results
from this type of work could be used to help prioritize communities and
species at particular risk to climate change impacts.

@@

Even the most complex marine model is a stylized cartoon of the true
dynamics at play in these systems. However, for all their limitations
models such as `marlin` can help users understand factors that drive the
performance of marine management strategies. The goal of this tool is to
empower people to design policies based on evaluation of key
uncertainties and trade-offs, and in doing so support more effective and
equitable marine resource management.

>>>>

if (!file.exists(here("data","sci_to_com.csv"))){
  sci_to_com <- taxize::sci2comm(unique(marlin_inputs$scientific_name),simplify= TRUE,db='ncbi') %>% 
  bind_rows(.id = "critter") %>% 
  pivot_longer(everything(),names_to = "critter",values_to = "common_name") %>% 
  mutate(common_name = stringr::str_to_title(common_name))

  write_csv(sci_to_com, file = here("data","sci_to_com.csv"))

} else{
 sci_to_com <-  read_csv(file = here("data","sci_to_com.csv"))

}

print(results_path)
results_files <- list.files(file.path(results_path))

@@

`marlin` follows a "models of intermediate complexity" philosophy
[@plagányi2014] targeted specifically towards capturing heterogeneous
habitats, movement dynamics, and the simultaneous impacts of fishing
fleets across multiple species (frequently called "technical
interactions"). In its general form, `marlin` simulates the behavior of
populations of spatially explicit biologically independent
age-structured animal populations affected by one or more fishing fleets
in time steps and a spatial resolution specified by the user.

While `marlin` does not simulate species interactions, fishing fleets in
`marlin` are capable of targeting and affecting multiple species
simultaneously. This allows the model to simulate processes such as
fisheries bycatch or effort displacement in a way that accounts for how
fleet behavior might affect multiple species in the system. The ability
to track the impacts of fishing fleets across multiple species
simultaneously is particularly important as very few fishing fleets are
truly single species.

As a demonstration of the use of this modeling tool, we explore two
general situations

1.  Trade-offs and distributional outcomes for food production and
    multi-species biomass in a coastal coral-reef fishery
2.  Implications of climate-driven range shifts for MPA design in a
    pelagic fishery

We chose these two examples to illustrate the use of `marlin` in
contrasting systems in which spatial management strategies such as MPAs are increasingly considered.

Below we provide a summary of the `marlin` model, as well as details of
the case studies. We note
that `marlin` is a complex software package with many options to control
detailed behaviors. Given that these specific features will evolve over
time, we do not describe in detail every option of the model here.
Rather, we focus on the equations and methods describing the core
foundations of our tool for efficiently simulating bio-economic dynamics
in aquatic systems.

While users can adapt the general equations and ideas presented here as
they see fit, the software package accompanying this paper,
[`marlin`](https://danovando.github.io/marlin/), is written in R
[@rcoreteam2021] and C++, integrated through the `Rcpp` package
[@eddelbuettel2018].

@@

`marlin` simulates the dynamics of one or more species, currently best
representing fish-like species, using age-structured population
dynamics. Each time step, species move throughout the simulated seascape
using both diffusion and "taxis" (active movement towards preferred
habitat), experience natural and potentially fishing mortality, and if
in a spawning season spawn using one of the possible forms of density
dependence implemented in the model, with the potential to include
auto-correlated stochastic deviations in the amount of offspring
produced, generally called "recruitment deviates".

These species can be caught by fishing fleets. A fishing fleet is defined
in `marlin` by a set of fishers that can have different fishing skill,
prices, and selectivity for individual species (each of which we denote
as a *metier*). For example, both a longline and purse seine fleet may
capture bigeye (*Thunnus obesus*) and skipjack tunas, but the longline
fleet can be made much more likely to catch bigeye than skipjack, and
*vice versa*. Each fishing fleet distributes its fishing effort in space
according to a specified spatial allocation rule, for example in
proportion to profit-per-unit-effort, conditional on management policies in places such as quotas and/or the presence of any spatial restrictions such as MPAs.

Unconstrained by management, the total amount of effort exerted by each
fleet can follow one of two dynamics: open access or constant effort.
Under constant effort, the total amount of effort of each fleet is fixed
over time, with the possible exception of attrition due to MPA
placement. Under open access, the total amount of effort in a given time
step is a function of the profitability of the fleet in the previous
time step, evolving until a bionomic equilibrium of zero total profits
is reached. Profitability is a function of the volume and price of each
species caught, as well as the cost of the total amount of fishing
effort per fleet and the travel costs as a function of distance from a port.

The dynamics of each fleet can be modified by management in a variety of
ways. Managers can impose size limits for individual species within each
fleet. They can also impose total catch quotas for each species in each
time step. When quotas are activated, if the total catch across all
fleets for a given species under the model's effort dynamics would
exceed the allowable quota for that species, the total effort for each
fleet is decreased proportional to its contribution to the total catch
until the quota is satisfied. As an alternative to catch quotas,
managers can set an effort cap per fleet, which prevents effort from
exceeding a given amount under open access dynamics, though effort may
be reduced below this cap if required by the profit equation (i.e. the
fleet can choose to fish less than the quota or effort cap).

`marlin` also allows users to specify closed fishing seasons for one or more
species in the system. During closed seasons, no effort is allocated to
patches that would result in capture of the covered species, potentially
resulting in zero effort if the species cannot be avoided. Under these
conditions, the amount of effort in the next open season is a function
of the effort and profits in the last open fishing season. Lastly,
managers can specify locations of no-take MPAs, which could in theory
change in size and location. When MPAs are implemented, fishing effort
that used to operate inside the MPAs can either leave the fishery, or be
redistributed outside the MPA (the default behavior).

Users can use the model to explore general properties of spatially
explicit social-ecological systems, or can parameterize it to explore
key uncertainties of more tactical applications. As an example of the
later case, we provide two case studies here designed to reflect the
kinds of questions stakeholders might face if designing an MPA network
for a coastal coral reef or a pelagic system.

@@

Revenues for each fleet in a time step are a function of the total
amount caught of each species and its price ($\Pi$). The amount
caught is a function of the selectivity at age of each species for each
fleet ($\alpha$), the fishing efficiency of the fleet for that that
species (also called "catchability", *q*), the amount of fishing effort
of the fleet in question in that patch in a given time step (*E*), and
the total fishing mortality at age (including all other fleets) for that
particular species in that patch and time (*u*).

@@

We approximate the dome-shaped form as a normal distribution with
mean $l^{sel}$ and standard deviation $\sigma$. The normal density
function is re-scaled such that the selectivity is 1 at $l^{sel}$.

@@

Users can tune their fleets in one of two ways. First, they can specify
a target exploitation rate *u* for each species in their simulation.
Taking all the other parameters of the model as given, `marlin` then
adjusts the catchability coefficients $q_{f,s}$ for each fleet *f* and
species *s* such that the equilibrium exploitation rate for each species
matches the desired level. Second, they can specify a target depletion,
defined as total spawning stock biomass under fishing divided by total
unfished spawning stock biomass, and the model will adjust the
catchability coefficients $q_{f,s}$ for each fleet *f* and species *s*
such that the equilibrium depletion for each species matches the desired
level.

@@

Each fishing fleet decides how to allocate its effort in space based on
one of four possible spatial allocation strategies. The ideal free
distribution (IFD) is the standard method for distributing fishing
fleets in spaces (see @gillis2003 and references therein). However,
analytical solutions to the IFD present a number of complications for
our model. Primarily, our model is capable of tracking multiple fleets
at once. This means that the IFD for a single fleet would commonly be modeled as a Nash-equilibrium based on the actions of each of the individual fleet
conditional on the actions of all other fleets. While possible to
implement, this would slow down our model runs to the point of
practically preventing large-scale evaluation of the spatial management policies.

@@

Revenue based spatial distribution is not likely to be very realistic;
in general we would expect fishing fleets to respond to profits on some
core level. However, due to the complexity of parameterizing cost
functions, fleet dynamics are often evaluated based on yield or revenue
alone, and so we include those scenarios here to allow users to evaluate
the potential implications of this choice. The decision on whether to
allocate the fleet based on absolute or relative (per unit effort)
metrics is more complex. When effort represents the actions of separate
and individual fishing actors (e.g. independent fishing vessels), a
per-unit-effort strategy, in which fishers distribute themselves based
on the expected catch of their individual efforts, may be more realistic
[@hilborn1987]. Conversely, a system defined by a sole owner seeking to
maximize total profits might be better represented by a fleet model
based on total profits

@@

The underlying population model used is an age structured single-species
model in the manner of @ovando2021a. The population model requires many
parameters. However, if the user supplies either a scientific or common
name for the species in question (scientific preferred), the model will
supply default values for that species based on the values reported in
FishLife [internet connection required, @thorson2020]. All of the life history values can
alternatively be replaced with user-supplied values as needed and should be for more tactical applications; the default values simply ensure that reasonable and internally consistent
values for the species in question are supplied as a starting point.

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of *advection* (follow currents), and *taxis* (active movement towards preferred habitat) and *diffusion* (essentially remaining variation in movement not explained by diffusion or taxis). For now, `marlin` focuses just on the
diffusion and taxis components of this model, though future extensions
could easily incorporate advection vectors from oceanographic models. In
this way, `marlin` allows users to run anything from a simple Gaussian
dispersal kernel up to a system governed by species that passively
diffuse around a dynamic thermal range.

We provide a brief overview of the the general CTMC method here, see
@thorson2021a for a detailed description. Under this framework, movement
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of the sum of the diffusion $\boldsymbol{D}$ and taxis $\boldsymbol{\tau}$ matrices scaled by the width of the time step $\Delta_{t}$ and the length of the edge of each cell $\Delta_d$ that the model is running under. So for example, if we are running a model on a monthly timestep, and we define the units of time in years, then $\Delta_{t} = 1/12$. Similarly, if we are running a model where one cell equals 100km^2^, then $\Delta_d = 100km$. This parameterization allows users to set the effective area of the spatial domain through two avenues; the number of cells, which effectively scales the resolution of the model, and the area of each cell, which scales the spatial extent of the simulation.

$$
\boldsymbol{M}_{t,s,a} = e^{\Delta_{t}(\boldsymbol{D}_{t,s,a} + \boldsymbol{\tau}_{t,s,a})} 
$$ {#eq-movement}

The individual components of the diffusion matrix are filled based on an adjacency matrix, which defines whether two cells are both adjacent and water (as opposed to land or another physical barrier), and a diffusion rate $D$ definend in units of length of the side of a cell per unit of time. For example, if we are defining the time units as years and the distance units as kilometers, for a tuna $D$ might be 1,000 $\frac{KM}{Year}$. The $\Delta_{t}$ and $\Delta_{p}$ parameters then translate the movement rate $D$ to match the time step and cell size used in a simulation. This "scale free" parameterization means that appropriate value of $D$ can be identified for a species and then set, regardless of the time step or cell size used in the simulation model itself.

$$ 
d_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}^2}D & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} d_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
\end{cases}
$$ {#eq-diffusion}

The taxis matrix $\boldsymbol{\tau}$ is built off of the difference in log habitat quality between each patch, the adjacency matrix, and temporal and spatial resolution of the model ($\Delta_{t}$ and $\Delta_{p}$), as well as a scalar term $\lambda$. $\lambda$ scales the magnitude of the habitat
gradient between patches relative to the diffusion
parameter $D$. When the diffusion and taxis matrices are estimated
jointly from actual data in the manner of @thorson2021a, $\lambda$
should be set to zero. However, when users are manually creating both
the diffusion and taxis matrices for simulation, the relative strength of diffusion to
taxis must be defined by the user. Setting higher values of $\lambda$
allows users to simulate a scenario where taxis is much more significant
than diffusion.

$$
\tau_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}}(1 + \lambda d_{s,f})e^{h(p2,t,s,a) - h(p1,t,s,a)} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} \tau_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
  \end{cases}
$$ {#eq-taxis}

For both the diffusion and taxis matrix, we allow for the inclusion of
physical barriers to movement (i.e. land). Pairs of patches that are
adjacent but in which one or both patches are a barrier to movement are
set as non-adjacent. The CTMC model then produces movement dynamics that
move around barriers rather than over them. This is different than
setting habitat in a patch to zero; an animal will not preferentially move
towards a patch with zero habitat all else being equal, but can move
over a zero habitat patch towards another better patch. They most move around a physical barrier.

@@

@@

We include two examples demonstrating how `marlin` can be used to
support marine spatial planning. In the first, we show how `marlin` can
be used to compare the total and distributional impacts of MPAs designed
in a heavily fished coastal coral reef ecosystem. In the second, we demonstrate
how `marlin` can be used to assess components of climate resilience of
alternative MPA design strategies in a pelagic system.

@@

Across all of our case studies we make use of two potential "rule of
thumb" MPA design strategies:

@@

2.  Target_fishing (*target_fishing*): MPAs locations are prioritized
    proportional to fishery catches. So, patches with high total
    catches are prioritized over patches with low catches.

@@

We use `marlin` to simulate the outcomes for both food production and
conservation for each of the species and both of the fleets as a
function of both MPA size and MPA design strategy. For this exercise,
MPAs are placed with perfect information and have no design constraints
for continuity (i.e. the selected MPA design can be as patchwork as
dictated by the data). We ran the simulation on a quarterly level, meaning that $\Delta_t = 1/ 4$ and set the area of each cell to an arbitrary value of 1 (meaning $\Delta_p = 1$ ).

@@

For this exercise, we focused on using `marlin` to assess resilience of
the selected "rule of thumb" MPA design strategies to a climate-driven
range shift. Specifically, we simulate an extreme example where the
centroid of each population shifts northward at a rate of 0.2 patches
per year, over a `r experiment_years` year time horizon (@fig-blue-ssb).
We designed MPA networks given the conditions in the starting year, and
then held that network constant over the years of the experiment,
running one simulation with and another without the climate-drive range
shift. We then compared the effects of this range shift on food
production and conservation outcomes from MPA networks designed based on
the pre-range shift world. We ran the simulation on a quarterly level, meaning that $\Delta_t = 1/4$ and set the area of each cell to an arbitrary value of 1 (meaning $\Delta_p = 1$ ).

@@

For the coastal coral reef example the *target_fishing* MPA design strategies was capable of producing a Pareto-improving, i.e. potentially win-win, outcome where the total
spawning biomass divided by unfished spawning biomass across all species
increase by up to 200% while maintaining at least the same levels of
fisheries yields as would be caught in the absence of any MPAs.
Conversely, achieving increases in total fish biomass through the *rate* strategy resulted in immediate losses in total yields, and in losses to both yields and total biomass in some cases
(@fig-coral-results).

These macro-level changes in total yield and conservation outcomes mask
some internal tradeoffs. Nearly all of the food production benefits from
this MPA network accrued to fleet one, while catches by fleet two were generally
negatively impacted under both scenarios. For example, the MPA sizes and designs that result in the largest increase in yields for fleet one would produce a near total
loss of fish catch for fleet two.

The *rate* MPA placement strategy produced
lower total fishery gains but minimized the negative impacts on fleet
two. However, this strategy produced much lower conservation gains
than the *target_fishing* strategy, only benefiting
deep snapper meaningfully, and in fact resulting in slight losses to the
nearshore snapper and grouper populations for MPA sizes less than roughly 75% of the
simulated space. The *target_fishing* strategy
resulted in rapid conservation gains across all of the simulated
species.

MPAs were capable of producing positive outcomes in terms of total
biomass and yields in the coastal coral reef example, but all the
tested strategies resulted in tradeoffs across at least one dimension. `marlin` could be further used
by a community such as the one simulated here to explore alternative design
strategies, or to assess options for distributing fishery benefits
across communities that may experience very different food provision outcomes from the
same MPA network.

```{r}
#| label: fig-coral-results
#| fig-cap: "Frontier in conservation and yield outcomes under different MPA design strategies (A) for coastal coral reef case study. Change in yield for each fleet (B) and depletion per species (C) as a function of MPA size and MPA design strategy for the coral reef case study."
#| fig-height: 4
#| fig-width: 6

@@

The yield impacts of the proposed MPA strategies were much more
sensitive to climate impacts for the purse seine fishery. Under the
*status quo* species distribution, the *target_fishing* MPA design
strategy produced increases in fisheries yields at sizes between 25-50%
of the simulated area; that same MPA network resulted in the near
complete loss of yields from the purse seine fleet when applied to the
climate change scenario. In contrast, the yield outcomes for the
longline fleet were relatively insensitive to the simulated range shifts
(@fig-blue-results).

From a conservation perspective, species such as albacore and swordfish
were relatively insensitive to the simulated range shifts, whereas
species such as bigeye tuna and silky shark showed extreme variation in
outcomes depending on the climate scenario. For bigeye tuna, under the
*status quo* conditions the *target_fishing* MPA design strategy
resulted in rapid increases in the population, but that same MPA network produced much smaller and slower gains under climate change.
However, for blue sharks this trend is inverted, with more rapid
conservation gains under the range shift scenario than the *status quo*
(@fig-blue-results).

@@

Marine ecosystems are driven by complex social-ecological dynamics.
Communities must often make decisions on how to manage these systems based on
limited empirical evidence. Modeling tools such as `marlin` can help
users answer important scientific questions and design marine management
policies informed by a better understanding of sensitivities to key
uncertainties.

@@

In this paper we have documented the core system of equations underlying
`marlin`, and demonstrated how `marlin` can be used to support efforts
such as MPA planning. The variation in both the magnitude and
distribution of costs and benefits stemming from alternative design
choices and states of nature shown in our case study results
demonstrates the value in our method. For example, our coastal coral
reef example illustrates both the potential for MPAs to benefit
conservation and food production in these systems, and the potential for
the same MPA to benefit some fleets and species while harming others when fishing
fleets affected by an MPA do not share the same objectives and species are not uniformly distributed. Our pelagic
case study illustrates how `marlin` can be used to assess the climate
resiliency of spatial management strategies, and identify strategies
that best meet the needs of both current and future conditions. Results
from this type of work could be used to help prioritize communities and
species at particular risk to climate change impacts.

@@

Even the most complex marine model is a stylized cartoon of the true
dynamics at play in the ocean. However, for all their limitations
models such as `marlin` can help users understand factors that drive the
performance of marine management strategies. The goal of this tool is to
empower people to design policies based on evaluation of key
uncertainties and trade-offs, and in doing so support more effective and
equitable marine resource management.
